Quantifying fish range shifts across poorly defined management boundaries

### 2021-10-11 Scripts/manuscript_acrossboundaries.Rmd
Inclued  missing intro paragraphs

In the United States (U.S.) North East (NE) Atlantic, many important fished species such as american lobster (*Homarus americanus*), yellowtail flounder (*Limanda ferruginea*), summer flounder (*Paralichthys dentatus*) and red hake (*Urophycis chuss*) have been shifting their distribution poleward  [@pinsky_fogarty_2012], and modelling exercises suggest that NE Atlantic species will shift will average over 600 km towards the end of the $21^{st}$ century under a high emission climate change scenario [@Morley:2018]. The mismatch between implemented management strategies and shifts in stocks across management jurisdictions compromises management and sustainability [@Baudron2020; @pinsky_fogarty_2012]. Fishers in the NE Atlantic have been adapting to such shifts by following the fish, in some cases, or changing target speices, in others [@polo_2021]. While this problem of shifting stocks across management jurisdictions is increasingly well known, quantifying it is difficult for two main reasons: (*i*) management jurisdictions in the ocean are often ill-defined, especially at the sub-national level or beyond Exclusive Economic Zones (EEZ), and (*ii*) quantifying fish distributions in relation to these boundaries is difficult, even where fishery-independent surveys exist, because of seasonal and interannual variation.

**[Short paragraph on the ill-defined boundary challeng - do we privilege a state's geographic claims or the inherent claims to the resource of existing fishing ports?]**

**[Short paragraph on the seasonality of fish distributions in temperate waters - when do we measure fish distributions?]**

@@

Although the stock-wide quota for our three target species is allocated among the states, U.S. Federal waters are not sub-divided into state-owned fisheries management regions (**hereafter refereed as regulatory units**). This lack of regulatory units within federal waters creates a challenge for assessing a stock's distribution with respect to individual states. There are a potentially infinite number of ways to geographically determine regulatory units with the EEZ's 200 nautical miles. Here we focus on two methods based which likely bracket much of the range of options.

@@

The first approach consists on expanding the current geographical distribution of States and their statutory state waters. For this approach we adapted the state's waters boundaries determined by the Federal and State Waters shapefile created for the Ocean Reporting Tool in April, 2018 ([NOAA's Office for Coastal Management, 2021](https://catalog.data.gov/dataset/federal-and-state-waters)). This shapefile consists in multiple polygons each one representing the statutory 3 miles waters that each state owns (Figure \@ref{fig:figureS1}). Each state's polygon  from Maine to Virginia was expanded to the 200 nm EEZ boundary using the *st* package in *R* [@REF]. The resulting polygons were also gridded into a 3 x 3 deg. matrix. The regions where the grid cells overlapped were equally allocated to the sharing states.

@@

The second approach to determine regulatory units gives greater weight to the historical landings of fishing ports. For this approach we used NOAA Fisheries landings-by-port data [@REF_NOAA]. Landings from each species and port were averaged from 2010 to 2019 noting that 2020-2021 were removed to avoid using COVID19-impacted years. We then estimated the total port-wide landings of the averaged period and the proportion that each port had in terms of landings *per* species. We selected the main ports landing each species, defined as those concentrating the top $75{^th}$ percentile of landings. Note that, while species often shared top ports, their final number and location varied according to each species (Figure \@ref{fig:figureS1}). Moreover, we tested the sensitivity of our results to the arbitrary $75{^th}$ percentile threshold and found that... (*to be done*). Each port represented a point in the map (i.e., latitude and longitude degrees) whose coordinates were manually collected using [googlemaps](www.googlemaps.com). We then drew a polygon from each port to the EEZ and gridded the expanded polygon following the same method as the *Fishing ports approach*. The expanded polygon of each port was associated to the state where the port is located and overlapping grids were equally allocated to the sharing states.

@@

Finally, we estimated the CPUE percentage change today relative to the historic period. For that matter we define the historic period as 1980 to 2000 representing the state allocations of the commercial fisheries originally implemented in 2003 [@RefDocument], and “today” as the average of 2009 to 2019 representing the more recent data available. The parentage changes between the historic and present CPUE ($\Delta{B}$) was estimate for each of the two regulatory units approaches as;

@@

Seasonal differences are also evident, although smaller than regulatory unit differences, when estimate each state's CPUE proportion for all species (Figures \@ref{fig:figSeasonlDiff}). Here, half of the cases represent 1% or no seasonality difference like Connecticut's CPUE proportion of summer flounder. The largest difference are in the latitudinal extremes with Maine currently getting 4% more CPUE in the fall than in spring as opposite to Virginia who gets 4% more in the spring than in the fall.

@@

The proportion of the CPUE has increased for northern states

>>>>

In the United States (U.S.) North East (NE) Atlantic, many important fished species such as american lobster (*Homarus americanus*), yellowtail flounder (*Limanda ferruginea*), summer flounder (*Paralichthys dentatus*) and red hake (*Urophycis chuss*) have been shifting their distribution poleward  [@pinsky_fogarty_2012], and modelling exercises suggest that NE Atlantic species will shift will average over 600 km towards the end of the $21^{st}$ century under a high emission climate change scenario [@Morley:2018]. The mismatch between implemented management strategies and shifts in stocks across management jurisdictions compromises management and sustainability [@Baudron2020; @pinsky_fogarty_2012]. Fishers in the NE Atlantic have been adapting to such shifts by following the fish, in some cases, or changing target species, in others [@polo_2021]. While this problem of shifting stocks across management jurisdictions is increasingly well known, quantifying it is difficult for two main reasons: (*i*) management jurisdictions in the ocean are often ill-defined, especially at the sub-national level or beyond Exclusive Economic Zones (EEZ), and (*ii*) quantifying fish distributions in relation to these boundaries is difficult, even where fishery-independent surveys exist, because of seasonal and interannual variation.

<!-- ## Management jurisdictions in the ocean are often ill-defined -->

<!-- **[Short paragraph on the ill-defined boundary challenge - do we privilege a state's geographic claims or the inherent claims to the resource of existing fishing ports?]** -->

<!-- Both defining and ill-defined management boundaries represent a challenge to fisheries management [@REF]. For example, territory dispute is among the most frequent sources of conflict between Asian countries [@spijkers_globalpatterns_2019], specially in the South China [@Beckman]. -->

Fisheries scientists and common-resource economist have long pledge for establishing property rights of a common-pull resource [@REF]. However, allocating such rights is not allays straightforward [@REF]. There are multiple ways to allocate property rights within country's EEZ. For example, regulatory units can be allocated considering fishing ports or communities as exemplified by many small scale fisheries across Latin America. In Mexico, spatial property rights are often given to fishing communities based on historical exploitation and investment in specific fisheries [@PalaciosAbrantes2018; @Basurto2005]. Moreover, the Chilean's government adopted the use of Territorial Use Rights for Fisheries (TURFs) for the management of benthic fisheries. Under this policy, fishers organizations are allocated a TURF based on port location and historical landings, among others [@González_2006; @gelcich_2010]. Within the U.S., state waters give clear fisheries rights to each state, federal waters (i.e, the 200 n, EEZ), on the other hand are not subdivided in sub-regulatory units, except for some specific cases. Pacific halibut (*Hippoglossus stenolepis*) managed by Canada and the United States from California to Alaska is based on regulatory units that extend along the countries' EEZs [@IPHC]. In this case, such regulatory units cross multiple states, like division 2A that goes from California to Washington [@REF].

<!-- **[Short paragraph on the seasonality of fish distributions in temperate waters - when do we measure fish distributions?]** -->

Seasonality of fish distributions in temperate waters is key for proper management of marine ressources [@Bell2014]. Along the U.S. Northeast EEZ, muny species like black sea bass (*Centropristis striata*) and scup (*Stenotomus chrysops*) present substantial latitudinal migration in response to water temperature fluctuation [@Bell2015; @Henderson2017]. Changes in distribution also follow a depth patter like the case of summer flounder (*Paralichthys dentatus*) who undergoes annual migrations from shallow coastal habitats in the fall to deeper habitats over winter (Terceiro, 2001; Sackett et al., 2007).  The consideration of such natural migration is key in the proper management of marine resources [@Bell2014].

@@

Although the stock-wide quota for our three target species is allocated among the states, U.S. Federal waters are not sub-divided into state-owned fisheries management regions (**hereafter refereed as regulatory units**). This lack of regulatory units within federal waters creates a challenge for assessing a stock's distribution with respect to individual states. There are a potentially infinite number of ways to geographically determine regulatory units with the EEZ's 200 nautical miles. Here we focus on two methods which likely bracket much of the range of options.

@@

The first approach consists of expanding the current geographical distribution of states and their statutory state waters. For this approach we adapted the state's waters boundaries determined by the Federal and State Waters shapefile created for the Ocean Reporting Tool in April, 2018 ([NOAA's Office for Coastal Management, 2021](https://catalog.data.gov/dataset/federal-and-state-waters)). This shapefile consists in multiple polygons each one representing the statutory 3 miles waters that each state owns (Figure \@ref{fig:figureS1}). Each state's polygon  from Maine to Virginia was expanded to the 200 nm EEZ boundary using the *st* package in *R* [@REF]. The resulting polygons (i.e., regulatory units) were also gridded into a 3 x 3 deg. matrix. The regions where the grid cells overlapped were equally allocated to the sharing states.

@@

The second approach to determine regulatory units gives greater weight to the historical landings of fishing ports. For this approach we used NOAA Fisheries landings-by-port data [@REF_NOAA]. Landings from each species and port were averaged from 2010 to 2019 noting that 2020-2021 were removed to avoid using COVID19-impacted years. We then estimated the total port-wide landings of the averaged period and the proportion that each port had in terms of landings *per* species. We selected the main ports landing each species, defined as those concentrating the top $75{^th}$ percentile of landings. Note that, while species often shared top ports, their final number and location varied according to each species (Figure \@ref{fig:figureS1}). Each port represented a point in the map (i.e., latitude and longitude degrees) whose coordinates were manually collected using [googlemaps](www.googlemaps.com). We then drew a polygon from each port to the EEZ and gridded the expanded polygon following the same method as the *Fishing ports approach*. The resulting regulatory unit of each port was associated to the state where the port is located and overlapping grids were equally allocated to the sharing states.

Moreover, we tested the sensitivity of our results to the arbitrary $75{^th}$ percentile threshold and found that... (*to be done*).

@@

Next, we estimated the CPUE percentage change today relative to the historic period. For that matter we define the historic period as 1980 to 2000 representing the state allocations of the commercial fisheries originally implemented in 2003 [@RefDocument], and “today” as the average of 2009 to 2019 representing the more recent data available. The parentage changes between the historic and present CPUE ($\Delta{B}$) was estimate for each of the two regulatory units approaches as;

@@

Seasonal differences in state’s CPUE proportion are also evident, although smaller than regulatory units differences (Figures \@ref{fig:figSeasonlDiff}). Here, half of the cases represent 1% or no seasonality difference like Connecticut's CPUE proportion of summer flounder. The largest difference are in the latitudinal extremes with Maine currently getting 4% more CPUE in the fall than in spring as opposite to Virginia who gets 4% more in the spring than in the fall.

@@

Over the last 40 years, the proportion of the CPUE has increased for northern states and decreased for southern states for all species (Figure 6). North Carolina and Virginia have been the most affected for all three species, regardless of the regulatory unit approach. Under the state waters approach, states from Rhode Island to Maine have increased their proportion of black sea bass and scup while the fishing ports approach favors Massachusetts, Rhode Island, and New York. States located “at middle” latitude do not seem to be affected by shifting distributions.  While the regulatory unit approach will completely remove some states (e.g., Maine) of the allocation beneficing others (e.g., New Jersey), it does not seem to affect historical trends. That is, gainers and losers in terms of historical trend will be the same, regardless of the approach.
